---
title: "Deciphering Parkinson's: Vocal Biomarkers and Demographic Insights"
author: "Fanny D'Souza"
date: "2023-11-29"
output:
  html_document: default
  pdf_document: default
---


```{r setup, include=FALSE}
options(repos = c(CRAN = "https://cran.rstudio.com/"))
library(ggplot2)

```




```{r}
# Set the working directory (if necessary)
setwd("C:/main/Parkinsons_VoiceMetrics")

# Read the CSV file
data <- read.csv("parkinsonsdata.csv")

# View the first few rows of the data
head(data)

```


```{r}
#Missing values by column
missing_values <- colSums(is.na(data))
print(missing_values)

```
```{r}
#Descriptive Statistics
summary(data)
```


```{r}
#Checking the distribution of voice metric: Jitter 

ggplot(data, aes(y = Jitter...)) +
  geom_boxplot() +
  labs(title="Box Plot of Jitter Abs", y="Jitter Values")

ggplot(data, aes(y = Jitter.Abs.)) +
  geom_boxplot() +
  labs(title="Box Plot of Jitter Abs", y="Jitter Values")

ggplot(data, aes(y = Jitter.RAP)) +
  geom_boxplot() +
  labs(title="Box Plot of Jitter RAP", y="Jitter Values")

ggplot(data, aes(y = Jitter.PPQ5)) +
  geom_boxplot() +
  labs(title="Box Plot of Jitter PPQ5", y="Jitter Values")

ggplot(data, aes(y = Jitter.DDP)) +
  geom_boxplot() +
  labs(title="Box Plot of Jitter DDP", y="Jitter Values")

```

```{r}

# Assuming 'data' is your dataframe and 'Jitter' is the column you're analyzing

# Remove rows where Jitter values are greater than 3e-4
clean_data <- data[data$Jitter <= 3e-4, ]

# Now 'clean_data' contains only the rows where Jitter values are less than or equal to 3e-4


```


```{r}
#Checking the distribution of voice metric: Shimmer

ggplot(data, aes(y = Shimmer)) +
  geom_boxplot() +
  labs(title="Box Plot of Shimmer", y="Shimmer Values")

ggplot(data, aes(y = Shimmer.dB.)) +
  geom_boxplot() +
  labs(title="Box Plot of Shimmer dB", y="Shimmer Values")

ggplot(data, aes(y = Shimmer.APQ3)) +
  geom_boxplot() +
  labs(title="Box Plot of Shimmer APQ3", y="Shimmer Values")

ggplot(data, aes(y = Shimmer.APQ5)) +
  geom_boxplot() +
  labs(title="Box Plot of Shimmer APQ5", y="Shimmer Values")

ggplot(data, aes(y = Shimmer.APQ11)) +
  geom_boxplot() +
  labs(title="Box Plot of Shimmer APQ11", y="Shimmer Values")

ggplot(data, aes(y = Shimmer.DDA)) +
  geom_boxplot() +
  labs(title="Box Plot of Shimmer DDA", y="Shimmer Values")


```


```{r}
#Checking the distribution of voice metric: NHR

ggplot(data, aes(y = NHR)) +
  geom_boxplot() +
  labs(title="Box Plot of NHR", y="NHR Values")


```


```{r}
#Checking the distribution of indicators of Parkinson's severity such as ‘motor_UPDRS’ and ‘total_UPDRS’

ggplot(data, aes(y = motor_UPDRS)) +
  geom_boxplot() +
  labs(title="Box Plot of motor_UPDRS", y="motor_UPDRS Values")


ggplot(data, aes(y = total_UPDRS)) +
  geom_boxplot() +
  labs(title="Box Plot of total_UPDRS", y="motor_UPDRS Values")


```
STATISTICAL ANALYSIS

Null Hypothesis:
1.	There is no relationship between age and voice metrics in Parkinson's patients.
Alternative Hypothesis:
1.	There is a significant relationship between age and voice metrics in Parkinson's patients.
```{r}
names(data)
```


```{r}
#Checking for normality in variables like age and voice metrics

#1.  'jitter...' 
hist(data$`Jitter...`, prob = TRUE, main = "Histogram of jitter...", xlab = "jitter... values", col = "lightblue")

# Overlay a normal distribution curve
curve(dnorm(x, mean = mean(data$`Jitter...`), sd = sd(data$`Jitter...`)), add = TRUE, col = "red", lwd = 2)


#2. Voice metric -a)Jitter.Abs.


# Create a histogram for the 'Jitter.Abs.' variable
hist(data$`Jitter.Abs.`, prob = TRUE, main = "Histogram of Jitter.Abs.", xlab = "Jitter.Abs. values", col = "lightblue")

# Overlay a normal distribution curve
curve(dnorm(x, mean = mean(data$`Jitter.Abs.`), sd = sd(data$`Jitter.Abs.`)), add = TRUE, col = "red", lwd = 2)



#3. Voice metric -a)Jitter.RAP

hist(data$`Jitter.RAP`, prob = TRUE, main = "Histogram of Jitter.RAP", xlab = "Jitter.RAP values", col = "lightblue")

curve(dnorm(x, mean = mean(data$`Jitter.RAP`, na.rm = TRUE), sd = sd(data$`Jitter.RAP`, na.rm = TRUE)), add = TRUE, col = "red", lwd = 2)



#4.Voice metric -a)Jitter.PPQ5

hist(data$`Jitter.PPQ5`, prob = TRUE, main = "Histogram of Jitter.PPQ5", xlab = "Jitter.PPQ5 values", col = "lightblue")

# Overlay a normal distribution curve
curve(dnorm(x, mean = mean(data$`Jitter.PPQ5`, na.rm = TRUE), sd = sd(data$`Jitter.PPQ5`, na.rm = TRUE)), add = TRUE, col = "red", lwd = 2)


#5. Voice metric -a)Jitter.DDP

hist(data$`Jitter.DDP`, prob = TRUE, main = "Histogram of Jitter.DDP", xlab = "Jitter.DDP values", col = "lightblue")

# Overlay a normal distribution curve
curve(dnorm(x, mean = mean(data$`Jitter.DDP`, na.rm = TRUE), sd = sd(data$`Jitter.DDP`, na.rm = TRUE)), add = TRUE, col = "red", lwd = 2)



#SHIMMER

#a)"Shimmer"

hist(data$Shimmer, prob = TRUE, main = "Histogram of Shimmer", xlab = "Shimmer values", col = "lightblue")

curve(dnorm(x, mean = mean(data$Shimmer, na.rm = TRUE), sd = sd(data$Shimmer, na.rm = TRUE)), add = TRUE, col = "red", lwd = 2)


#b)Shimmer.dB

hist(data$`Shimmer.dB.`, prob = TRUE, main = "Histogram of Shimmer.dB.", xlab = "Shimmer.dB. values", col = "lightblue")

curve(dnorm(x, mean = mean(data$`Shimmer.dB.`, na.rm = TRUE), sd = sd(data$`Shimmer.dB.`, na.rm = TRUE)), add = TRUE, col = "red", lwd = 2)


#c)Shimmer.APQ3

hist(data$Shimmer.APQ3, prob = TRUE, main = "Histogram of Shimmer.APQ3", xlab = "Shimmer.APQ3 values", col = "lightblue")

curve(dnorm(x, mean = mean(data$Shimmer.APQ3, na.rm = TRUE), sd = sd(data$Shimmer.APQ3, na.rm = TRUE)), add = TRUE, col = "red", lwd = 2)


#d)Shimmer.APQ5

hist(data$Shimmer.APQ5, prob = TRUE, main = "Histogram of Shimmer.APQ5", xlab = "Shimmer.APQ5 values", col = "lightblue")

curve(dnorm(x, mean = mean(data$Shimmer.APQ5, na.rm = TRUE), sd = sd(data$Shimmer.APQ5, na.rm = TRUE)), add = TRUE, col = "red", lwd = 2)


#e)Shimmer.APQ11

hist(data$Shimmer.APQ11, prob = TRUE, main = "Histogram of Shimmer.APQ11", xlab = "Shimmer.APQ11 values", col = "lightblue")

curve(dnorm(x, mean = mean(data$Shimmer.APQ11, na.rm = TRUE), sd = sd(data$Shimmer.APQ11, na.rm = TRUE)), add = TRUE, col = "red", lwd = 2)



#f)Shimmer.DDA

hist(data$Shimmer.DDA, prob = TRUE, main = "Histogram of Shimmer.DDA", xlab = "Shimmer.DDA values", col = "lightblue")

curve(dnorm(x, mean = mean(data$Shimmer.DDA, na.rm = TRUE), sd = sd(data$Shimmer.DDA, na.rm = TRUE)), add = TRUE, col = "red", lwd = 2)


#NHR


hist(data$NHR, prob = TRUE, main = "Histogram of NHR", xlab = "NHR values", col = "lightblue")

curve(dnorm(x, mean = mean(data$NHR, na.rm = TRUE), sd = sd(data$NHR, na.rm = TRUE)), add = TRUE, col = "red", lwd = 2)



```
```{r}
#CORRELATION BETWEEN AGE AND JITTER

correlation_age_jitter <- cor.test(data$age, data$`Jitter.Abs.`, method = "spearman")
print(correlation_age_jitter)

#Similar for the rest, but check the next code for a better approach


```
```{r}
#Spear man's correlation between age and voice metrics

voice_metrics <- c("Jitter.Abs.", "Jitter.RAP", "Jitter.PPQ5", "Jitter.DDP", 
                   "Shimmer", "Shimmer.dB.", "Shimmer.APQ3", "Shimmer.APQ5", 
                   "Shimmer.APQ11", "Shimmer.DDA", "NHR","Jitter...")

alpha <- 0.05

# Create an empty data frame to store results
results <- data.frame(Variable = character(0), 
                       Correlation = numeric(0), 
                       p_value = numeric(0), 
                       Significant = character(0))

# Perform correlation tests for each variable
for (variable in voice_metrics) {
  correlation_result <- cor.test(data$age, data[[variable]], method = "spearman")
  is_significant <- ifelse(correlation_result$p.value < alpha, "Yes", "No")
  results <- rbind(results, data.frame(Variable = variable, 
                                       Correlation = correlation_result$estimate, 
                                       p_value = correlation_result$p.value, 
                                       Significant = is_significant))
}

# Print the results
print(results)

```
Null: The mean values of voice metrics are the same for both sexes.
Alternate:	The mean values of voice metrics differ between the sexes.

```{r}
#Mann Whitney Test to check for significant differences between 2 independent groups(Male and Female) and  voice metrics


voice_metrics <- c("Jitter.Abs.", "Jitter.RAP", "Jitter.PPQ5", "Jitter.DDP",  "Shimmer","Shimmer.dB.","Shimmer.APQ3","Shimmer.APQ5","Shimmer.APQ11", "Shimmer.DDA"  
,"NHR")

alpha <- 0.05

# Create an empty data frame to store results
results <- data.frame(Variable = character(0), p_value = numeric(0), Significant = character(0))

# Perform Mann-Whitney U tests for each variable
for (variable in voice_metrics) {
  mwu_test <- wilcox.test(data[[variable]] ~ data$sex, na.action = "na.omit")
  is_significant <- ifelse(mwu_test$p.value < alpha, "Yes", "No")
  results <- rbind(results, data.frame(Variable = variable, p_value = mwu_test$p.value, Significant = is_significant))
}

print(results)


```
	RQ2: Can voice metrics, along with demographic factors (age and sex), reliably predict the severity of Parkinson's disease as indicated by UPDRS scores?

Null Hypothesis:
1.	Voice metrics, age, and sex cannot predict the severity of Parkinson's disease.
Alternative Hypothesis:
1.	Voice metrics, age, and sex can significantly predict the severity of Parkinson's disease.

```{r}

names(data)

```

```{r}

model <- lm(total_UPDRS ~ age + sex + Jitter... + Jitter.Abs. + Jitter.RAP + Jitter.PPQ5 + Jitter.DDP + Shimmer + Shimmer.dB., data = data)

```

```{r}

# Check linearity
plot(model, which = 1)

# Check independence of errors
plot(model, which = 3)

# Check homoscedasticity
plot(model, which = 2)

# Check normality of residuals
hist(residuals(model))
qqnorm(residuals(model))
qqline(residuals(model))

```
```{r}
summary(model)
```

